---
title: "Preparing flow-data for use with with `beadplexr`"
author: "Ulrik Stervbo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Preparing flow-data for use with with beadplexr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Introduction

Because of potential license problems with some dependencies in the `flowCore`-package, the functionality of reading and preparing flow-data for `beadplexr` is removed. This vignette described how to quickly achieve the same functionality as before the issues.

The following will not be executed when the vignette is built, but should work when executed manually, providing all packages are installed.

## Reading FACS data

Install `flowCore` and its dependencies:

```{r inst-flow-core, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("flowCore")
```

Source the code found in section [Copy-paste code].

The output of a LEGENDplex experiment is a series of Flow Cytometry Standard (FCS) files. The exact version of the fcs-files produced, depends on the flow cytometer used. However, getting data into R should be the same irrespective of the cytometer used.

The **beadplexr** provides the function `read_fcs()` to read in a fcs file. It is a wrapper around the functionality provided by the [**flowCore**](http://bioconductor.org/packages/flowCore/) package and performs the following steps:

  1. Read the file with `flowCore::read.FCS()`
  2. Apply an *arcsinh* transformation of the APC and PE channels 
  3. Remove boundary events of of the forward and side scatter channels
  4. Optionally subset the channels to contain just bead events -- this might improve identification of the beads
  5. Convert the FACS data to a `data.frame`
  
```{r read-fcs-defaults, eval=FALSE}
library(beadplexr)

.file_name <-
  system.file("extdata", "K2-C07-A7.fcs", package = "beadplexr")

# Load the fcs file with all defaults
df <- read_fcs(.file_name = .file_name)
```

The **beadplexr** package also comes with some simple plot functions to help visualize the FACS data. The functions are wrappers around [**ggplot2**](https://ggplot2.tidyverse.org/), and are meant to provide an easy view of the data. For more control it is probably better to use **ggplot2** directly.

```{r raw-events, eval=FALSE, fig.show='hold'}
facs_plot(df)
facs_plot(df, .x = "FL2-H", .y = "FL6-H")
```

## Applying the event filter
In the above example we used the default filter settings. These settings fit nicely with the settings of our flow cytometer, but might not work so well for yours, so the following demonstrates the arguments to the `read_fcs()`.

Here we apply no filter at all:
```{r read-fcs-no-filter, eval=FALSE, fig.show='hold', eval=FALSE}
df <- read_fcs(.file_name = .file_name, .filter = NULL)

facs_plot(df)
facs_plot(df, .x = "FL2-H", .y = "FL6-H")
```

And here we look at the beads with the larger forward and side scatter, but using the height parameter in stead of the area parameter:
```{r fcs-h-filter, eval=FALSE, fig.show='hold'}
df <- read_fcs(.file_name = .file_name, .fsc_ssc = c("FSC-H", "SSC-H"), 
                  .filter = list("FSC-H" = c(3.75e5L, 5.5e5L),
                                 "SSC-H" = c(4e5, 1e6),
                                 "FL6-H" = c(7L, Inf)))

facs_plot(df, .x = "FSC-H", .y = "SSC-H")
facs_plot(df, .x = "FL2-H", .y = "FL6-H")
```

If you insist, you can also filter the PE channel, but that is probably not such a good idea:

```{r read-fcs-fl2-filter, eval=FALSE, fig.show='hold', eval=FALSE}
df <- read_fcs(.file_name = .file_name, .fsc_ssc = c("FSC-H", "SSC-H"), 
                  .filter = list("FSC-H" = c(3.75e5L, 5.5e5L),
                                 "SSC-H" = c(4e5, 1e6),
                                 "FL6-H" = c(7L, Inf),
                                 "FL2-H" = c(8, 10)))

facs_plot(df, .x = "FSC-H", .y = "SSC-H")
facs_plot(df, .x = "FL2-H", .y = "FL6-H")
```

## Pseudo-color plots

If you prefer, and have **hexbin** installed, you can view the FACS data with the commonly used pseudo-color.

```{r pseudo-color, eval=FALSE, fig.show='hold'}
df <- read_fcs(.file_name = .file_name, .fsc_ssc = c("FSC-A", "SSC-A"),
                  .filter = list("FSC-A" = c(4e5, 6e5L),
                                 "SSC-A" = c(4e5, 1e6),
                                 "FL6-H" = c(7L, Inf)))

# We have relatively few events (around 2500), so decreasing the number of bins
# make the scale more visible
facs_plot(df, .type = "hexbin", .bins = 50)
facs_plot(df, .x = "FL2-H", .y = "FL6-H", .type = "hexbin")
```


## Copy-paste code

Place the code below in a file and source the R-script.

```{r eval=FALSE}
#' Read a fcs file.
#'
#' This is a wrapper around reading a FACS data file using \code{flowCore},
#' transforming the channels with bead signal events using
#' \code{arcsinhTransform()} from \code{flowCore}, automatic removing boundary
#' events of the forward and side scatter channels, subsetting channels (if
#' needed), and cast to a \code{data.frame}.
#'
#' @param .file_name The path and name of the file to be read.
#' @param .fsc_ssc The names of the forward and side scatter channels. A
#'   character vector of length of two.
#' @param .bead_channels The names of the channels with bead events. A character
#'   vector of length of at least two.
#' @param .filter Optional list of upper and lower cut-off for individual
#'   channels. Use \code{.filter = NULL} for no filtering at all.
#' @param .compensation A character vector, a compensation matrix, or
#'   \code{NULL}. See 'Details' for extended information of the argument.
#'
#' @param ... additional arguments passed to \code{flowCore::read.FCS}
#'
#' @details
#'
#' The \code{.compensation} argument takes a character vector or an actual
#' compensation matrix. In case of the latter, it must be a object of class
#' \code{matrix}. If an object of class \code{character} is given to the
#' \code{.compensation} argument, it can be the keyword 'guess' or a search
#' pattern matching the keyword in the fcs file that contains the compensation
#' matrix. If the keyword 'guess' is passed, the function looks for a matrix at
#' the keywords "SPILL" and "SPILLOVER", and if none are found it takes the
#' first matrix which contains the parameter names as column names. Finally, the
#' values of \code{.compensation} can also be \code{NULL}, in which case, no
#' compensation is performed.
#'
#' To summarize, the argument \code{.compensation} can be:
#'
#' \describe{
#'   \item{A matrix}{The compensation matrix to apply}
#'   \item{A character}{
#'     \describe{
#'       \item{The word 'guess'}{Looks for a matrix at the keywords "SPILL"
#'       and "SPILLOVER". If none found the first matrix with parameter names is
#'       applied}
#'       \item{A search string}{The first matrix found within the keywords
#'       matching the given search string is applied. The search string can be a
#'       regular expression}
#'     }
#'   }
#'   \item{NULL}{Nothing is done}
#' }
#'
#' @return A \code{data.frame} with the columns given in \code{.fsc_ssc} and \code{.bead_channels}.
#' @export
#'
#' @seealso \code{\link[flowCore]{rectangleGate}} for instructions to the
#'   \code{.filter} list argument, \code{\link[flowCore]{boundaryFilter}}
#'   for automatic removal of boundary events, and
#'   \code{\link[flowCore]{arcsinhTransform}} for the transformation.
#'
#'
#' @examples
#' library(beadplexr)
#'
#' .file_name <- system.file("extdata", "K2-C07-A7.fcs",
#'                           package = "beadplexr")
#'
#' # Load the fcs file, with no filter
#' df <- read_fcs(.file_name = .file_name, .filter = NULL)
#'
#' plot(df[, c("FSC-A", "SSC-A")])
#' plot(df[, c("FL2-H", "FL6-H")])
#'
#' # Load the fcs file, with default filter
#' df <- read_fcs(.file_name = .file_name, .fsc_ssc = c("FSC-H", "SSC-H"))
#'
#' plot(df[, c("FSC-H", "SSC-H")])
#' plot(df[, c("FL2-H", "FL6-H")])
#'
#' # Load the fcs file, with custom filter
#' df <- read_fcs(.file_name = .file_name,
#'                   .filter = list("FSC-A" = c(2e5L, 3.3e5L),
#'                                  "SSC-A" = c(2e5, 1e6L),
#'                                  "FL2-H" = c(11L, 14L),
#'                                  "FL6-H" = c(9L, Inf)))
#'
#' plot(df[, c("FSC-A", "SSC-A")])
#' plot(df[, c("FL2-H", "FL6-H")])
#'
#' # Specify three bead channels
#'
#' df <- read_fcs(.file_name = .file_name,
#'                   .bead_channels = c("FL6-H", "FL2-H", "FL1-H"))
#'
#' plot(df[, c("FSC-A", "SSC-A")])
#' plot(df[, c("FL2-H", "FL6-H")])
#' plot(df[, c("FL1-H", "FL2-H")])
#' plot(df[, c("FL1-H", "FL6-H")])
#'
read_fcs <- function(.file_name, .fsc_ssc = c("FSC-A", "SSC-A"), .bead_channels = c("FL6-H", "FL2-H"),
                     .filter = list("FSC-A" = c(2e5L, 8e5L), "SSC-A"= c(2e5L, 1e6L), "FL6-H" = c(7.3, Inf)),
                     .compensation = "guess", ...){
  
  if(!is.null(.fsc_ssc)){
    if(length(.fsc_ssc) != 2 | !is.character(.fsc_ssc)){
      stop(".fsc_ssc must be a character vector of length 2")
    }
  }
  
  if(!is.null(.bead_channels)){
    if(length(.bead_channels) < 2 | !is.character(.bead_channels)){
      stop(".bead_channels must be a character vector of at least 2")
    }
  }
  
  
  flowCore::read.FCS(filename = .file_name, transformation = FALSE, ...) %>%
    transform_bead_channels(.bead_channels = .bead_channels) %>%
    apply_compensation(.compensation = .compensation) %>%
    remove_boundary_events(.channels = c(.fsc_ssc)) %>%
    subset_channels(.filter = .filter) %>%
    as_data_frame_flow_frame(.channels = c(.fsc_ssc, .bead_channels))
}

#' Remove boundary events.
#'
#' A wrapper around application of \code{flowCore}'s \code{boundaryFilter}.
#'
#' @param .flow_frame A \code{flowFrame}. Usually the result of \code{read.FCS} from \code{flowCore}.
#' @param .channels A character vector with the channels to apply the filter to (mostly just forward and side scatter).
#'
#' @return A \code{FlowFrame} with border events removed.
#'
#' @seealso \code{\link[flowCore]{boundaryFilter}} for comments on the filter.
#'
#' @keywords internal
#'
#' @examples
#' \dontrun{
#' library(beadplexr)
#' library(flowCore)
#'
#' .file_name <- system.file("extdata", "K2-C07-A7.fcs",
#'                           package = "beadplexr")
#'  # Load the fcs file
#'  .flow_frame <- read.FCS(filename = .file_name,
#'                          transformation = FALSE)
#'
#' # Plot with all events
#' w_bdr <- .flow_frame@exprs
#' plot(w_bdr[, c("FSC-A", "SSC-A")])
#'
#' # Remove boundary events
#' .flow_frame <- remove_boundary_events(.flow_frame,
#'                           .channels = c("FSC-A", "SSC-A"))
#'
#' wo_bdr <- as.data.frame(.flow_frame@exprs)
#' plot(wo_bdr[, c("FSC-A", "SSC-A")])
#' }
remove_boundary_events <- function(.flow_frame, .channels){
  if(is.null(.channels)){
    return(.flow_frame)
  }
  
  bf <- flowCore::boundaryFilter(x = .channels, side = "both", filterId = "Automatic Boundary")
  bf <- flowCore::filter(x = .flow_frame, filter = bf)
  flowCore::Subset(.flow_frame, bf)
}

#' Subset channels
#'
#' A wrapper around the application of \code{flowCore}'s \code{rectangleGate}.
#'
#' @param .flow_frame A \code{flowFrame}. Usually the result of \code{read.FCS}
#'   from \code{flowCore}.
#' @param .filter An Optional list of upper and lower cutoff for individual
#'   channels. Default is no filtering at all.
#'
#' @return A \code{FlowFrame} with border events removed.
#'
#' @seealso \code{\link[flowCore]{rectangleGate}} for instructions to the
#'   \code{.filter} list argument
#'
#' @keywords internal
#'
#' @examples
#' \dontrun{
#' library(beadplexr)
#' library(flowCore)
#'
#' .file_name <- system.file("extdata", "K2-C07-A7.fcs",
#'                             package = "beadplexr")
#' # Load the fcs file
#'  .flow_frame <- read.FCS(filename = .file_name,
#'                           transformation = FALSE)
#'
#'  # Plot with all events
#'  all_events <- .flow_frame@exprs
#'  plot(all_events[, c("FSC-A", "SSC-A")])
#'  # Events are untransformed
#'  plot(all_events[, c("FL2-H", "FL6-H")])
#'
#'  # A silly thing that does nothing
#'  .flow_frame <- subset_channels(.flow_frame)
#'  filter_1 <- .flow_frame@exprs
#'  plot(filter_1[, c("FSC-A", "SSC-A")])
#'
#'  # Filter on forward scatter
#'  .flow_frame <- subset_channels(.flow_frame,
#'                                 .filter = list("FSC-A" = c(1e3L, 2e6)))
#'  filter_2 <- .flow_frame@exprs
#'  plot(filter_2[, c("FSC-A", "SSC-A")])
#'
#'  # Filter on forward and side scatter
#'  .flow_frame <- subset_channels(.flow_frame,
#'                                 .filter = list("FSC-A" = c(2e5L, 6.5e5L),
#'                                                 "SSC-A" = c(2e5, 1e6L)))
#'  filter_3 <- .flow_frame@exprs
#'  plot(filter_3[, c("FSC-A", "SSC-A")])
#'
#'  # Filter on the unfiltered bead channels (please transform first)
#'  .flow_frame <- subset_channels(.flow_frame,
#'                                 .filter = list("FL2-H" = c(5e4L, 6e5L),
#'                                                "FL6-H" = c(0, 3e5L)))
#'  filter_4 <- .flow_frame@exprs
#'  plot(filter_4[, c("FL2-H", "FL6-H")])
#'
#'  # And everything at once
#'  .flow_frame <- read.FCS(filename = .file_name,
#'                         transformation = FALSE)
#' .flow_frame <- subset_channels(.flow_frame,
#'                                .filter = list("FSC-A" = c(2e5L, 6.5e5L),
#'                                               "SSC-A" = c(2e5, 1e6L),
#'                                               "FL2-H" = c(5e4L, 6e5L),
#'                                               "FL6-H" = c(0, 3e5L)))
#' filter_5 <- .flow_frame@exprs
#' plot(filter_5[, c("FSC-A", "SSC-A")])
#' plot(filter_5[, c("FL2-H", "FL6-H")])
#' }
subset_channels <- function(.flow_frame, .filter = NULL){
  if(is.null(.filter)){
    return(.flow_frame)
  }
  
  gf <- flowCore::rectangleGate(.filter, filterId = "Manual Boundary")
  gf <- flowCore::filter(x = .flow_frame, filter = gf)
  flowCore::Subset(.flow_frame, gf)
}

#' Cast \code{flowFrame} to \code{data_frame}.
#'
#' Extracts the flow events from the \code{flowFrame} as a \code{data_frame}
#'
#' @param .flow_frame A \code{flowFrame}. Usually the result of \code{read.FCS} from \code{flowCore}.
#' @param .channels A character vector with the event channels to extract. Default is all channels.
#'
#' @return A \code{data_frame}
#' @keywords internal
#'
#' @examples
#' \dontrun{
#' library(beadplexr)
#' library(flowCore)
#' .file_name <- system.file("extdata", "K2-C07-A7.fcs",
#'                           package = "beadplexr")
#' # Load the fcs file
#' .flow_frame <- read.FCS(filename = .file_name,
#'                         transformation = FALSE)
#'
#' # Get all channels
#' as_data_frame_flow_frame(.flow_frame)
#'
#' # Just interesting
#' as_data_frame_flow_frame(.flow_frame,
#'                          .channels = c("FSC-A", "SSC-A",
#'                                        "FL2-H", "FL6-H"))
#' }
as_data_frame_flow_frame <- function(.flow_frame, .channels = NULL){
  if(is.null(.channels)){
    .channels <- flowCore::featureNames(.flow_frame)
  }
  
  .flow_frame@exprs[, .channels] %>%
    as.data.frame.matrix()
    #tibble::as_tibble()
}

#' Transform parameters with bead events.
#'
#' Channels with bead events are transformed using \code{flowCore}'s \code{arcsinhTransform()}.
#'
#' @param .flow_frame A \code{flowFrame}. Usually the result of \code{read.FCS} from \code{flowCore}.
#' @param .bead_channels A character vector of the length of 2 with the names of the channels to transform.
#'
#' @return A \code{flowFrame} with the beads channels transformed
#'
#' @keywords internal
#'
#' @seealso \code{\link[flowCore]{arcsinhTransform}} for the transformation.
#'
#' @examples
#' \dontrun{
#' library(beadplexr)
#' library(flowCore)
#' .file_name <- system.file("extdata", "K2-C07-A7.fcs",
#'                           package = "beadplexr")
#' # Load the fcs file
#' .flow_frame <- read.FCS(filename = .file_name,
#'                                  transformation = FALSE)
#' # Transform channels
#' .flow_frame <- transform_bead_channels(.flow_frame = .flow_frame,
#'                             .bead_channels =
#'                                 c("FL6-H", "FL2-H"))
#' }
transform_bead_channels <- function(.flow_frame, .bead_channels){
  if(is.null(.bead_channels)){
    return(.flow_frame)
  }
  if(length(.bead_channels) < 2){
    stop(".bead_channels must at least be of length 2")
  }
  # If any of the values in bead_channels are not found we cannot transform or
  # do anything
  in_fn <- .bead_channels %in% flowCore::featureNames(.flow_frame)
  
  if(FALSE %in% in_fn){
    stop(paste(paste(.bead_channels[!in_fn], collapse = " and "), "not found in the flow frame", sep = " "))
  }
  
  # Transform channels and return
  transformtion_list <- flowCore::transformList(.bead_channels, flowCore::arcsinhTransform(), transformationId = "defaultArcsinhTransform")
  flowCore::transform(.flow_frame, transformtion_list)
}

#' Apply compensation
#'
#' A compensation matrix is applied
#'
#' @param .flow_frame A \code{flowFrame}. Usually the result of \code{read.FCS}
#'   from \code{flowCore}.
#' @param .compensation A character vector, a compensation matrix, or
#'   \code{NULL}. See 'Details' for extended information of the argument.
#'
#' @details
#'
#' The \code{.compensation} argument takes a character vector or an actual
#' compensation matrix. In case of the latter, it must be a object of class
#' \code{matrix}. If an object of class \code{character} is given to the
#' \code{.compensation} argument, it can be the keyword 'guess' or a search
#' pattern matching the keyword in the fcs file that contains the compensation
#' matrix. If the keyword 'guess' is passed, the function looks for a matrix at
#' the keywords "SPILL" and "SPILLOVER", and if none are found it takes the
#' first matrix which contains the parameter names as column names. Finally, the
#' values of \code{.compensation} can also be \code{NULL}, in which case, no
#' compensation is performed.
#'
#' To summarize, the argument \code{.compensation} can be:
#'
#' \describe{
#'   \item{A numerical matrix}{The compensation matrix to apply}
#'   \item{A character}{
#'     \describe{
#'       \item{The word 'guess'}{Looks for a matrix at the keywords "SPILL"
#'       and "SPILLOVER". If none found the first matrix with parameter names is
#'       applied}
#'       \item{A search string}{The first matrix found within the keywords
#'       matching the given search string is applied. The search string can be a
#'       regular expression}
#'     }
#'   }
#'   \item{NULL}{Nothing is done}
#' }
#'
#' @return A compensated \code{flowFrame}
#'
#' @keywords internal
#'
#'
#' @seealso \code{\link[flowCore]{compensation-class}} for the compensation.
#'
#' @examples
#' \dontrun{
#' library(beadplexr)
#' library(flowCore)
#' .file_name <- system.file("extdata", "K2-C07-A7.fcs",
#'                           package = "beadplexr")
#' # Load the fcs file
#' .flow_frame <- read.FCS(filename = .file_name,
#'                                  transformation = FALSE)
#' # Apply compensation by guessing the matrix
#' .flow_frame <- apply_compensation(.flow_frame = .flow_frame,
#'                             .compensation = "guess")
#' }
#'
apply_compensation <- function(.flow_frame, .compensation){
  # Got nothing, do nothing
  if(is.null(.compensation)){
    return(.flow_frame)
  }
  
  if(is.character(.compensation) & length(.compensation) > 1){
    warning(".compensation can only have the length of 1. I am using just the first element")
    .compensation <- .compensation[1]
  }
  
  # Test if .x is a numerical matrix with the parameters
  .get_potential_comp_matrices <- function(.x, .p){
    if(("matrix" %in% class(.x)) & is.numeric(.x)){
      TRUE %in% (.p %in% colnames(.x))
    }else{
      FALSE
    }
  }
  
  # Search keywords and unify warning
  .search_for_keywords <- function(.keywords, .pattern){
    .matches <- grepl(.pattern, .keywords, ignore.case = TRUE) %>% which
    if(length(.matches) > 1){
      warning("Found more than one potential compensation matrix. Applying the first.")
      .matches <- .matches[1]
    }
    .matches
  }
  
  # We get either a character or something which can be turned into a numerical matrix
  if(is.character(.compensation)){
    ff_keywords <- flowCore::keyword(.flow_frame)
    ff_parameters <- flowCore::colnames(.flow_frame)
    
    # No matter if we guess or were given a search word, we need a numerical
    # matrix where the parameters are in the column names. Otherwise, the actual
    # compensation application will fail.
    num_matrix <- ff_keywords %>%
      purrr::map_lgl(.get_potential_comp_matrices, .p = ff_parameters)
    num_matrix <- ff_keywords[num_matrix]
    
    # If we found nothing we cannot do nothing - except give a warning
    if(length(num_matrix) == 0){
      warning("No compensation applied. I could not find an appropriate matrix in the flow frame.")
      return(.flow_frame)
    }
    
    if(.compensation == "guess"){
      # Do the numerical matricies have names that we could expect?
      match_names <- names(num_matrix) %>% .search_for_keywords(.pattern = "SPILL")
      
      if(length(match_names) > 0){
        comp_matrix <- num_matrix[[match_names]]
      }else{
        # We found no standard keywords. If we have more than one potential
        # matrix, we warn
        if(length(num_matrix) > 1){
          warning("Found more than one potential compensation matrix. Applying the first.")
        }
        comp_matrix <- num_matrix[[1]]
      }
    }else{
      # A search we have
      match_names <- names(num_matrix) %>% .search_for_keywords(.pattern = .compensation)
      if(length(match_names) > 0){
        comp_matrix <- num_matrix[[match_names]]
      }else{
        warning("Your search word gave no match. No compensation applied")
        return(.flow_frame)
      }
      
    }
  }
  else{
    if(is.numeric(.compensation)){
      comp_matrix <- .compensation
    }else{
      stop("I don't know how to deal with what you gave me. Please check the documentation.")
    }
  }
  
  flowCore::compensate(.flow_frame, comp_matrix)
}
```
